Expression QTL hotspot dissection with mediation analysis and pleiotropy testing

### 2019-06-24 analysis/paper/paper.Rnw
minor fixes to paper text

For each hotspot, we performed a series of mediation analyses and pleiotropy
tests. For each hotspot, we identified a set of local expression traits and a
set of nonlocal expression traits. All traits demonstrated LOD peaks that met
the genome-wide significance threshold. The nonlocal expression traits arose from genes
on other chromosomes, while the local traits reflected expression levels for
genes on the same chromosome as the hotspot and within 7 Mb of the hotspot
center.

@@

For each hotspot, we paired each local trait with each nonlocal trait and performed a
two-dimensional, two-QTL scan over a genomic region that included the entirety of the
hotspot. We fitted bivariate linear mixed effects models at each ordered pair of markers \citep{boehm2019testing}.

@@

From the two-dimensional scans, we calculated pleiotropy test statistics for each pair of traits \citep{jiang1995multiple,boehm2019testing}.

\begin{equation}
\Lambda = \max_{\lambda_1, \lambda_2}ll(\lambda_1, \lambda_2) - \max_{\lambda} ll(\lambda, \lambda)
\label{eq:lrt}
\end{equation}

@@

For each hotspot and each local trait - nonlocal trait pair, we also performed a
mediation analysis. Each mediation analysis involved fitting four linear models (Equations~\ref{model1}, \ref{model2}, \ref{model3}, \ref{model4}).

@@

In all mediation analyses, the nonlocal transcripts are the targets, while the local transcripts serve as candidate mediators of the QTL-target relationship.

>>>>

For each hotspot, we identified a set of local expression traits and a
set of nonlocal expression traits. All traits demonstrated LOD peaks that met
the genome-wide significance threshold, 7.18 \citep{keller2018genetic}.
The nonlocal expression traits arose from genes
on other chromosomes, while the local traits reflected expression levels for
genes on the same chromosome as the hotspot and within 4 Mb of the hotspot
center. We ignored transcripts for genes on the hotspot chromosome but not within 4 Mb of
the hotspot center.

@@

For each hotspot, we paired each local trait with each nonlocal trait and performed a
two-dimensional, two-QTL scan over a genomic region that included the entirety of the
hotspot. We fitted bivariate linear mixed effects models at each ordered pair of markers in a two-dimensional grid \citep{boehm2019testing}.

@@

From the two-dimensional scans, we calculated pleiotropy likelihood ratio test statistics for each
pair of traits \citep{jiang1995multiple,boehm2019testing, boehm2019thesis}. Because of the required
computing time for bootstrap analyses, we didn't obtain p-values for the pleiotropy tests. Instead,
we worked directly with the test statistic values.

\todo[inline]{copy equations from G3 paper}

\begin{equation}
LOD = \log_{10} \left(\frac{\max_{\lambda_1, \lambda_2}L(B, \Sigma, \lambda_1, \lambda_2)}{\max_{\lambda}L(B, \Sigma, \lambda, \lambda)} \right)
\label{eq:lrt}
\end{equation}

\begin{equation}
ll(\lambda_1, \lambda_2)
\end{equation}

@@

For each hotspot and each local trait - nonlocal trait pair, we also performed a
mediation analysis to assess the extent which the local trait mediates the QTL effect on the nonlocal trait. Each mediation analysis involved fitting four linear models (Equations~\ref{model1}, \ref{model2}, \ref{model3}, \ref{model4}) \citep{chick2016defining}.

@@

In all mediation analyses, the nonlocal transcripts are the targets, while the local
transcripts serve as candidate mediators of the QTL-target relationship.
In Equations~\ref{model1} through \ref{model4}, $X$ is a matrix of founder allele dosages,
$B$ is a matrix
of founder allele effects, $W$ is a matrix of covariates, $C$ is a matrix of covariate
effects, $M$ is a vector of values for a single candidate mediator, $\beta$ is the effect
of the mediator on the complex trait, 1 is a vector with all entries set to 1, $\alpha$
is the mean trait value, and $E$ is a vector of normally distributed random errors. \todo[inline]{better so say 'column matrix' or 'vector'? I hate to worry about this.}
